\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-25 coqimpl.tex

\textbf{Family compiled into module.} This is the main idea of compiling
family components into Coq's primitives: we compile families into Coq
modules, family types into module types, and the context of a given term
into parameters of the module. Unlike module, we need to faithfully
reflect the \textit{late binding} nature of families: for each defining
field $ .. \cL\sigma \vdash t : T $, we need to compile $t$ using
``universal-quantifier-wrapped term'' : $.. \vdash \lambda t : \forall
\cL \sigma. T$. However, instead of using Coq's universal quantifier, we
choose module and module parameters to achieve this wrapping---each
family type in the context of the judgement will be one module parameter
with the compiled module type, and thus each field of the family will
compile into a functor parametrized by its context. Doing so we can also
get rapid feedback from type checking when the users are defining each
field interactively with Coq.
% Insert one pseudo-code example for explanation of the mechanism
\begin{figure}[!htb]
  \begin{minipage}[t]{0.25\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family A.
Field a : nat 
  := 1.
Field b 
  : a = a
  := eq_refl.
Final Field a' : nat 
  := 1. 
Field b'
  : a' = 1
  := eq_refl.
EndFamily.
\end{minted}
  \end{minipage}\YZ{Can the self's be omitted?}\EDJreply{Yes. But in our current plugin implementation it is not omitted. I think altering example is fine. If you think some small distance between example and the current implementation is fine, please comment back and I will remove all the self_}\YZreply{If omitting the self's poses no technical challenges, then I'd say it's fine to also omit them in this figure.}\EDJreply{Omitting self_ is easy and not much engineering effort is required. But omitting self_ will lead to `A.a' instead of `a' actually. So if we really want `a' as you last time mentioned, then a resolution is required and more engineering effort is required. But I don't see any technical difficulties. I have removed all the self, please check.}
  \begin{minipage}[t]{0.4\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module a_4 (self_A: EmptySig).
Definition a : nat := 1. End a_4.
Module Type a_5 (self_A: EmptySig).
Parameter (a : nat). End a_5.
Module Type a_6 := a_5 EmptyMod.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module b_7 (self_A: a_6).
Definition b 
  : self_A.a = self_A.a 
  := eq_refl. End b_7.

@@

% Concrete details of compilation of a family term data structure
Taking \cref{fig:plugin-example1} as an example: when the user
interactively inputs the line "Field a : nat := ..", our plugin will
translate this statement into the "Module a_4", "Module Type a_5, a_6"
three components.
{The "EmptySig" is just empty module type.} Then,
these modules and module types will be generated and type-checked by Coq
immediately. The compiled "Module a_4"\YZ{should this be a_4?}\EDJreply{Yes. Thanks!} is then part of the context of
"Field b" (as the module type of "self_A") for "b" to refer to. The
internal representation of family "A" is the list of these
compiled modules and module types. We can achieve incremental
checking and overriding as in \ref{langdesign-req0}---for example, if we
want to override "Field a", we simply replace that compiled module with
the new module inside the list. This compiled module type serves as an
abstraction that enables overriding.

Now that we know a family data structure is actually a list of compiled
modules, we may want to aggregate them into one complete compiled
module. We achieve that by repeatedly using Coq's "Include" command as
the example illustrated. Note that, in our example, we include two
functors without arguments---this is due to a perculiarity of Coq's
"Include": non-instantiated parameters will be automatically
instantiated by the appropriate fields in the surrounding defining
module.\YZ{'For example, ...'}\EDJreply{I don't think explaining a Coq feature inside our paper is a good idea? But I have added this in the following line, please check. Keep it and resolve the comment if you find this ok. }
For example, "b_7" requires a module of type "a_6". Inside module "A", when "Include a_4" is done, we will have a field \mintinline{Coq}{a : nat} inside the current surrounding "A". Then when we \mintinline{Coq}{Include b_7} without specifying the module parameters, Coq will try to find fields in the surrounding module, (i.e. module "A") to instantiate \mintinline{Coq}{self_A: a_6}. This is satisfied by the \mintinline{Coq}{a : nat} that is included earlier.
The resulting compiled module is just like mundane Coq's module and thus maintains all the computational mechanism from Coq, fulfilling \ref{langdesign-req0b}.

% explaining inheritance
Besides standalone family, we also need to implement family inheritance. To simplify the implementation, we can consider all families as inheritance and the standalone families are extending empty family. Thus we only need to deal with inheritance during interactive proving.

An inheritance data will be annotated with the type for the input family
and output family, and generally there are three kinds of
inheritance---extend, override and inherit. Extending corresponds to
adding new fields, overriding corresponds to modifying existent fields,
and inheriting corresponds to directly taking fields from the parent.
Thus, it is possible to consider inheritance as solely a data rather than
a functor, and this actually provides us the possibility to
\text{mix-in} inheritance data.

\textbf{Reasoning Requires Non-Overridable.} This compilation strategy
can allow fields to be type-dependent on former fields thus fulfilling
\ref{langdesign-req1}. However, there are some cases we need to expose the concrete definition of a given fields (especially when we are proving properties about it).

For example, we may
define $"add" : \mathbb{Z} \to \mathbb{Z} \to \mathbb{Z}$ as one field
and intend to prove its commutativity in a following field $"comm"$.
But to prove so, we need to know more than just a type interface of
"add" but also its \textbf{concrete fixpoint definition}. Otherwise, in the
future, once "add" is overridden into a subtraction function then
"comm" will have problem to be inherited.
A more concrete example is in \cref{fig:plugin-example1}, consider the
case we have "Field b : self_A.a = 1 := .." instead. Then our
compilation and Coq's type checking will both fail: in the context of
"b" (module type "a_6") we only know "a" is of type "nat" but nothing
more (and this is necessary so that "a" can be overridden in the
future).

To fix that, we need more information on "a" than its type---we
need to expose the concrete definition "a = 1" to the context of "b".
\YZ{I had a hard time working through these two paras. I guess they are
saying that there is a need to expose definitional equalities, and
therefore we need non-overridables. This causal flow should be made
more explicit.}\EDJreply{Good suggestion. I try to add this casual flow into the first line of the paragraph. Please check and resolve if find satisfied.}

In our Coq plugin, we can achieve such exposure of definition by
exposing the whole definition directly into the compiled module type. Doing so we also prohibit overriding because
subtraction doesn't have the same definition as $add$. We require the
users to decorate field as `Final' so that the plugin will proceed with
this special treatment.\YZ{This para is confusing: which kind of equality is used, propositional or judgmental? Where does 'Final' show up in the example?}\EDJreply{I rephrase it and remove the part on taking about meta-theory. Now everything is about judgemental so I don't have to emphasize judgemental. I wanted to make the example small but I guess I can add one `Final' statement into our example. Don't resolve it until I add it into Figure 2.}

>>>>

\textbf{Family compiled into module.} This is the main idea of compiling
family components into Coq's primitives: we compile families into Coq
modules, family types into module types, and the context of a given term
into parameters of the module. Unlike module, we need to faithfully
reflect the \textit{late binding} nature of families: for each defining
field $ .. \cL\sigma \vdash t : T $, we need to compile $t$ using
``universal-quantifier-wrapped term'' : $.. \vdash \lambda t : \forall
\cL \sigma. T$. However, instead of using Coq's universal quantifier, we
choose module and module parameters to achieve this wrapping---each
family type in the context of the judgement will be one module parameter
with the compiled module type, and thus each field of the family will
compile into a functor parametrized by its context. Doing so we can also
get rapid feedback from type checking when the users are defining each
field interactively with Coq.
% Insert one pseudo-code example for explanation of the mechanism
\begin{figure}[!htb]
  \begin{minipage}[t]{0.25\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family A.
Field a : nat 
  := 1.
Field b : a = a
  := eq_refl.
Final Field a' : nat 
  := 1. 
Field b' : a' = 1
  := eq_refl.
EndFamily.
\end{minted}
  \end{minipage}%\YZ{Can the self's be omitted?}\EDJreply{Yes. But in our current plugin implementation it is not omitted. I think altering example is fine. If you think some small distance between example and the current implementation is fine, please comment back and I will remove all the self_}\YZreply{If omitting the self's poses no technical challenges, then I'd say it's fine to also omit them in this figure.}\EDJreply{Omitting self_ is easy and not much engineering effort is required. But omitting self_ will lead to `A.a' instead of `a' actually. So if we really want `a' as you last time mentioned, then a resolution is required and more engineering effort is required. But I don't see any technical difficulties. I have removed all the self, please check.}
  \begin{minipage}[t]{0.4\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module a_4 (self_A: EmptySig).
Definition a : nat := 1. End a_4.
Module Type a_5 (self_A: EmptySig).
Parameter (a : nat). End a_5.
Module Type a_6 := a_5 EmptyMod.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module b_7 (self_A: a_6).
Definition b 
  : self_A.a = self_A.a 
  := eq_refl. End b_7.

@@

% Concrete details of compilation of a family term data structure
Taking \cref{fig:plugin-example1} as an example: when the user
interactively inputs the line "Field a : nat := ..", our plugin will
translate this statement into the "Module a_4", "Module Type a_5, a_6"
three components.
{The "EmptySig" is just empty module type.} Then,
these modules and module types will be generated and type-checked by Coq
immediately. The compiled "Module a_4" is then part of the context of
"Field b" (as the module type of "self_A") for "b" to refer to. The
internal representation of family "A" is the list of these
compiled modules and module types. We can achieve incremental
checking and overriding as in \ref{langdesign-req0}---for example, if we
want to override "Field a", we simply replace that compiled module with
the new module inside the list. This compiled module type serves as an
abstraction that enables overriding.

Now that we know a family declaration is compiled to a list of Coq
modules and module types, we may want to aggregate them into one complete compiled
module.\YZ{The reader wonders why this aggregation is needed. Seems like the place to talk about how the programmer can use A outside Family A.}
We achieve that by repeatedly using Coq's "Include" command as
the example illustrates. Note that, in our example, we include two
functors without arguments---this is due to a peculiarity of Coq's
"Include": non-instantiated parameters will be automatically
instantiated by the appropriate fields in the surrounding defining
module.
For example, "b_7" requires a module of type "a_6". Inside module "A",
when "Include a_4" is done, we will have a field \mintinline{Coq}{a : nat}
inside the current surrounding "A". Then when we
\mintinline{Coq}{Include b_7} without specifying the module parameter,
Coq will try to find fields in the surrounding module (i.e., module
"A") to instantiate \mintinline{Coq}{self_A: a_6}. This is satisfied by
the \mintinline{Coq}{a : nat} that is included earlier.
The resulting compiled module is just like mundane Coq's module and thus maintains all the computational mechanism from Coq, fulfilling \ref{langdesign-req0b}.

% explaining inheritance
Besides standalone family, we also need to implement family inheritance.
To simplify the implementation, we can consider all families as
inheritance and the standalone families are extending the empty family.
Thus we need to deal with only inheritance during interactive proving.

An inheritance data\YZ{what is 'inheritance data'?} will be annotated with the type for the input family
and output family. There are three kinds of inheritance---extend,
override, and inherit. Extending corresponds to adding new fields,
overriding corresponds to modifying existent fields, and inheriting
corresponds to directly taking fields from the parent.
Thus, it is possible to consider inheritance as solely a data rather
than a functor, and this actually provides us the possibility to
\text{mix-in} inheritance data.

\textbf{Reasoning Requires Non-Overridable.} This compilation strategy
can allow fields to be type-dependent on former fields thus fulfilling
\ref{langdesign-req1}. However, there are cases where we need to expose
the concrete definition of a given field (especially when we are proving
properties about it).

For example, we may
define $"add" : \mathbb{Z} \to \mathbb{Z} \to \mathbb{Z}$ as one field
and intend to prove its commutativity in a following field $"comm"$.
But to prove so, we need to know more than just a type interface of
"add" but also its \textbf{concrete fixpoint definition}. Otherwise, in the
future, once "add" is overridden into a subtraction function then
"comm" will have problem to be inherited.\YZ{The sentence reads as if you want to enable overriding "add", whereas what you really want is to expose the definition equality about "add".}
A more concrete example is in \cref{fig:plugin-example1}, consider the
case we have "Field b : a = 1 := .." instead. Then our
compilation and Coq's type checking will both fail: in the context of
"b" (module type "a_6") we only know "a" is of type "nat" but nothing
more (and this is necessary so that "a" can be overridden in the
future).\YZ{Not clear if you want to be able to override a or want to be able to have 'Field b : a = 1'.}

To fix that, we need more information on "a" than its type---we
need to expose the concrete definition "a = 1" to the context of "b".
%\YZ{I had a hard time working through these two paras. I guess they are
%saying that there is a need to expose definitional equalities, and
%therefore we need non-overridables. This causal flow should be made
%more explicit.}\EDJreply{Good suggestion. I try to add this casual flow into the first line of the paragraph. Please check and resolve if find satisfied.}

In our Coq plugin, we can achieve such exposure of definitional equality by
exposing the whole definition directly into the compiled module type. Doing so we also prohibit overriding because
subtraction doesn't have the same definition as $add$. We require the
users to decorate field as `Final' so that the plugin will proceed with
this special treatment.\YZ{This para is confusing: which kind of equality is used, propositional or judgmental? Where does 'Final' show up in the example?}\EDJreply{I rephrase it and remove the part on taking about meta-theory. Now everything is about judgemental so I don't have to emphasize judgemental. I wanted to make the example small but I guess I can add one `Final' statement into our example. Don't resolve it until I add it into Figure 2.}
